An Efficient Minibatch Acceptance Test for Metropolis-Hastings

### 2016-10-04 nips_2016.tex
updates

\begin{proof}
The following bound is given immediately after Corollary 2 in from~\cite{explicit-clt05}:
\begin{equation}\label{eq:clt-boundsx}
-6.4E|X|^3-2E|X| \leq    \sup_x|{\rm Pr}(t<x) - \Phi(x)|{\sqrt{n}} \leq 1.36E|X|^3
\end{equation}
This bound applies to $x\geq 0$. Applying the bound to $-x$ when $x<0$
and combining with $x>0$, we obtain the weaker but unqualified bound
in Equation~\ref{eq:clt-bounds}.
\end{proof}

@@

\begin{itemize}[noitemsep]
    \item The proof of Lemma~\ref{lem:worst_case}.
    \item More information on Section~\ref{ssec:gaussians}.
\end{itemize}

@@

Assume $(\theta' - \theta) \in \pm\frac{1}{\sqrt{N}}[0.5,1]$ and $\theta -0.5
\in \pm\frac{1}{\sqrt{N}}[0.5,1]$ with matching sign (i.e., first establish
$(\theta'-\theta)$, then ``force'' $\theta-0.5$ to be the same sign).  As
discussed in Lemma~\ref{lem:worst_case}, both $q(\theta' \mid \theta)$ and
$p(\theta \mid x_1,\ldots,x_N)$ have variance $1/N$. If we denote $\Phi$ as the
CDF of the standard normal distribution, then the former event occurs with
probability $p_0 = 2(\Phi(\sqrt{N}\frac{1}{\sqrt{N}}) -
\Phi(\sqrt{N}\frac{0.5}{\sqrt{N}})) = 2(\Phi(1)-\Phi(0.5)) \approx 0.2997$. The
latter event, because we restrict signs, occurs with probability $p_1 =
\Phi(\sqrt{N}\frac{1}{\sqrt{N}}) - \Phi(\sqrt{N}\frac{0.5}{\sqrt{N}}) = \Phi(1)
- \Phi(0.5) \approx 0.14988$.

These events together guarantee that $\Lambda^*(\theta,\theta')$ is negative,
intuitively because the next choice of $\theta'$ must be ``further away'' from
0.5 than $\theta$.  This fact means that we can find a $u \in (0,1)$ so that
$\psi(u,\theta,\theta') = \log u < 0$ equals $E[\Lambda^*(\theta,\theta')]$.
Specifically, choose $u_0$ to satisfy $\log u_0 = E[\Lambda^*(\theta,\theta')]$.
Using $E[x_i] = 0.5$ and Equation~\ref{eq:lemma_ll_ratio}, we see that
\begin{align}
    \log u_0 &= N(\theta'-\theta)\frac{1}{b} \cdot E\left[\sum_{i=1}^b x_i-\theta-\frac{\theta'-\theta}{2}\right] \\
    &= -N(\theta'-\theta)\left(\theta-0.5+\frac{\theta'-\theta}{2}\right).
\end{align}
Next, consider the minibatch acceptance test $\Lambda^*(\theta,\theta') \not\approx
\psi(u,\theta,\theta')$ required by these methods, where $\not\approx$
informally means ``significantly different from'' under the distribution over
samples of $x_i$. This turns out to be
\begin{align}
\Lambda^*(\theta,\theta') \hspace{0.1in} \not\approx \hspace{0.1in}
\psi(u_0,\theta,\theta') &\iff N(\theta'-\theta) \cdot \frac{1}{b}\sum_{i=1}^b x_i-\theta-\frac{\theta'-\theta}{2} \hspace{0.1in} \not\approx \hspace{0.1in} \log u_0 \\
&\iff \frac{1}{b}\sum_{i=1}^b x_i-\left(\theta+\frac{\theta'-\theta}{2} + \frac{\log u_0}{N(\theta'-\theta)}\right) \hspace{0.1in} \not\approx \hspace{0.1in} 0 \\
&\iff \frac{1}{b}\sum_{i=1}^b x_i-0.5 \hspace{0.1in} \not\approx \hspace{0.1in} 0. \label{eq:accept_test_zero}
\end{align}
Since the $x_i$ have mean 0.5, the resulting test with our chosen $u_0$ will
never correctly succeed and must use all $N$ data points.  Furthermore, if we
sample values of $u$ near enough to $u_0$, the terms in parenthesis will not be
sufficiently different from 0.5 to allow the test to succeed.

@@

\section{Gaussian Mixture Experiment Details}\label{app:gaussian}

In this section, we provide more details on Section~\ref{ssec:gaussians}.  Our
parameter is a 2-D vector
$\theta = \langle \theta_1, \theta_2 \rangle$, where
\begin{equation}
    \theta_1 \sim \mathcal{N}(0, \sigma_1^2) \quad \mbox{ and } \quad \theta_2 \sim \mathcal{N}(0, \sigma_2^2)
\end{equation}
and where $\mathcal{N}$ indicates the normal distribution (more generally, the
multivariate normal). We consider the above as our prior.
Following~\cite{langevin_2011}, we set $\sigma_1^2 = 10$ and $\sigma_2^2=1$, so
the covariance matrix of $\theta$ is $\Sigma = {\rm diag}(10,1)$. Therefore, the
log prior probability we endow on $\theta$ is
\begin{equation}
    \log p(\theta) = \log \left(\frac{1}{2\pi\sqrt{10}}\right) - \frac{1}{2}\theta^T\Sigma^{-1}\theta.
\end{equation}
To generate the data, we use the following Gaussian mixture with tied means:
\begin{equation}\label{eq:x_points}
    x_i \sim \frac{1}{2}\mathcal{N}(\theta_1, \sigma_x^2) + \frac{1}{2}\mathcal{N}(\theta_1+\theta_2, \sigma_x^2)
\end{equation}
where, again following~\cite{langevin_2011}, we set $\sigma_x^2 = 2$ and we draw
the data with $\theta_1=0$ and $\theta_2=1$ fixed (though these values are
arbitrary, and different choices would yield different posterior shapes).
This means the log likelihood of a single data instance is
\begin{equation}\label{eq:log_ll_single_instance}
    \log p(x_i \mid \theta) = \log \frac{1}{4\sqrt{\pi}} +
    \log\left(\exp\left(-\frac{1}{4}(x_i - \theta_1)^2\right) + \exp\left(-\frac{1}{4}(x_i -
    (\theta_1+\theta_2))^2\right)\right)
\end{equation}
The official problem statement is as follows: given some number of conditionally
independent data points $x_1, x_2, \ldots, x_N$ generated according
to~(\ref{eq:x_points}), determine the posterior distribution of $\theta$:
\begin{equation}\label{eq:log_post}
    \log p(\theta \mid x_1,\ldots,x_N) = \log p(\theta) + \sum_{i=1}^N\log p(x_i \mid \theta).
\end{equation}
Alternatively, depending on the application, we may opt to instead pick a point
estimate\footnote{We do this for the logistic regression experiment, described
in Section~\ref{ssec:logistic}.} of $\theta$, such as the MAP estimate. (If $N$
is extremely large, it will cause the posterior to peak sharply at its modes,
reducing distribution estimates to point estimates.)

One technique we use to make our test more applicable is to add a temperature to
the posterior. In general, we will add $T > 1$ so that the posterior is
$p(\theta)(\prod_{i=1}^N p(x_i\mid \theta))^{1/T}$, resulting in the log
posterior of 
\begin{equation}\label{eq:log_prior_temp}
    \log p(\theta \mid x_1,\ldots,x_N) \approx \log p(\theta) + \frac{1}{T}\sum_{i=1}^N\log p(x_i \mid \theta).
\end{equation}
which has the extra $1/T$ to decrease the scale factor, so the variance scales
as $1/T^2$.

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=1\linewidth]{contour_v2}
  \caption{The posterior distribution, with $N=10000$ and temperature $T$ varying.}
  \label{fig:contour2}
\end{figure}

To gain intuition on the shape of the posterior, Figure~\ref{fig:contour2} shows
simulated contour plots of the posterior based on $N=10000$ data points, and
temperature varying.  The (non-tempered) posterior is a multimodal distribution
with modes at $(0,1)$ and $(1,-1)$. A larger $T$ implies a flatter posterior,
which can be useful for parameter exploration so that the samples do not get
trapped in a mode.

>>>>

\begin{proof}
The following bound is given immediately after Corollary 2 from~\cite{explicit-clt05}:
\begin{equation}\label{eq:clt-boundsx}
-6.4E|X|^3-2E|X| \leq    \sup_x|{\rm Pr}(t<x) - \Phi(x)|{\sqrt{n}} \leq 1.36E|X|^3
\end{equation}
This bound applies to $x\geq 0$. Applying the bound to $-x$ when $x<0$
and combining with $x>0$, we obtain the weaker but unqualified bound
in Equation~\ref{eq:clt-bounds}.
\end{proof}

@@

\begin{itemize}[noitemsep]
    \item The proof of Lemma~\ref{lem:worst_case}.
\end{itemize}

@@

Choose $(\theta' - \theta) \in \pm\frac{1}{\sqrt{N}}[0.5,1]$ (event 1) and $(\theta -0.5)
\in \pm\frac{1}{\sqrt{N}}[0.5,1]$ filtered for  matching sign (event 2).  As
discussed in Lemma~\ref{lem:worst_case}, both $q(\theta' \mid \theta)$ and
$p(\theta \mid x_1,\ldots,x_N)$ have variance $1/N$. If we denote $\Phi$ as the
CDF of the standard normal distribution, then the former event occurs with
probability $p_0 = 2(\Phi(\sqrt{N}\frac{1}{\sqrt{N}}) -
\Phi(\sqrt{N}\frac{0.5}{\sqrt{N}})) = 2(\Phi(1)-\Phi(0.5)) \approx 0.2997$. The
latter event, because we restrict signs, occurs with probability $p_1 =
\Phi(1) - \Phi(0.5) \approx 0.14988$.

These events together guarantee that $\Lambda^*(\theta,\theta')$ is negative
by inspection of equation (\ref{eq:whyneg}) below.
This implies that we can find a $u \in (0,1)$ so that
$\psi(u,\theta,\theta') = \log u < 0$ equals $E[\Lambda^*(\theta,\theta')]$.
Specifically, choose $u_0$ to satisfy $\log u_0 = E[\Lambda^*(\theta,\theta')]$.
Using $E[x_i] = 0.5$ and Equation~\ref{eq:lemma_ll_ratio}, we see that
\begin{equation}
  \log u_0 = N(\theta'-\theta)\frac{1}{b} \cdot E\left[\sum_{i=1}^b x_i-\theta-\frac{\theta'-\theta}{2}\right]
\end{equation}
\begin{equation}\label{eq:whyneg}
    log u_0 = -N(\theta'-\theta)\left(\theta-0.5+\frac{\theta'-\theta}{2}\right).
\end{equation}
Next, consider the minibatch acceptance test $\Lambda^*(\theta,\theta') \not\approx
\psi(u,\theta,\theta')$ used in ~\cite{cutting_mh_2014} and~\cite{icml2014c1_bardenet14} , where $\not\approx$
means ``significantly different from'' under the distribution over
samples of $x_i$. This turns out to be
\begin{align}
\Lambda^*(\theta,\theta') \hspace{0.1in} \not\approx \hspace{0.1in}
\psi(u_0,\theta,\theta') &\iff N(\theta'-\theta) \cdot \frac{1}{b}\sum_{i=1}^b x_i-\theta-\frac{\theta'-\theta}{2} \hspace{0.1in} \not\approx \hspace{0.1in} \log u_0 \\
&\iff \frac{1}{b}\sum_{i=1}^b x_i-\left(\theta+\frac{\theta'-\theta}{2} + \frac{\log u_0}{N(\theta'-\theta)}\right) \hspace{0.1in} \not\approx \hspace{0.1in} 0 \\
&\iff \frac{1}{b}\sum_{i=1}^b x_i-0.5 \hspace{0.1in} \not\approx \hspace{0.1in} 0. \label{eq:accept_test_zero}
\end{align}
Since the $x_i$ have mean 0.5, the resulting test with our chosen $u_0$ will
never correctly succeed and must use all $N$ data points.  Furthermore, if we
sample values of $u$ near enough to $u_0$, the terms in parenthesis will not be
sufficiently different from 0.5 to allow the test to succeed.

@@

%% \section{Gaussian Mixture Experiment Details}\label{app:gaussian}

%% In this section, we provide more details on Section~\ref{ssec:gaussians}.  Our
%% parameter is a 2-D vector
%% $\theta = \langle \theta_1, \theta_2 \rangle$, where
%% \begin{equation}
%%     \theta_1 \sim \mathcal{N}(0, \sigma_1^2) \quad \mbox{ and } \quad \theta_2 \sim \mathcal{N}(0, \sigma_2^2)
%% \end{equation}
%% and where $\mathcal{N}$ indicates the normal distribution (more generally, the
%% multivariate normal). We consider the above as our prior.
%% Following~\cite{langevin_2011}, we set $\sigma_1^2 = 10$ and $\sigma_2^2=1$, so
%% the covariance matrix of $\theta$ is $\Sigma = {\rm diag}(10,1)$. Therefore, the
%% log prior probability we endow on $\theta$ is
%% \begin{equation}
%%     \log p(\theta) = \log \left(\frac{1}{2\pi\sqrt{10}}\right) - \frac{1}{2}\theta^T\Sigma^{-1}\theta.
%% \end{equation}
%% To generate the data, we use the following Gaussian mixture with tied means:
%% \begin{equation}\label{eq:x_points}
%%     x_i \sim \frac{1}{2}\mathcal{N}(\theta_1, \sigma_x^2) + \frac{1}{2}\mathcal{N}(\theta_1+\theta_2, \sigma_x^2)
%% \end{equation}
%% where, again following~\cite{langevin_2011}, we set $\sigma_x^2 = 2$ and we draw
%% the data with $\theta_1=0$ and $\theta_2=1$ fixed (though these values are
%% arbitrary, and different choices would yield different posterior shapes).
%% This means the log likelihood of a single data instance is
%% \begin{equation}\label{eq:log_ll_single_instance}
%%     \log p(x_i \mid \theta) = \log \frac{1}{4\sqrt{\pi}} +
%%     \log\left(\exp\left(-\frac{1}{4}(x_i - \theta_1)^2\right) + \exp\left(-\frac{1}{4}(x_i -
%%     (\theta_1+\theta_2))^2\right)\right)
%% \end{equation}
%% The problem statement is as follows: given some number of conditionally
%% independent data points $x_1, x_2, \ldots, x_N$ generated according
%% to~(\ref{eq:x_points}), determine the posterior distribution of $\theta$:
%% \begin{equation}\label{eq:log_post}
%%     \log p(\theta \mid x_1,\ldots,x_N) = \log p(\theta) + \sum_{i=1}^N\log p(x_i \mid \theta).
%% \end{equation}
%% Alternatively, depending on the application, we may opt to instead pick a point
%% estimate\footnote{We do this for the logistic regression experiment, described
%% in Section~\ref{ssec:logistic}.} of $\theta$, such as the MAP estimate. (If $N$
%% is extremely large, it will cause the posterior to peak sharply at its modes,
%% reducing distribution estimates to point estimates.)

%% One technique we use to make our test more applicable is to add a temperature to
%% the posterior. In general, we will add $T > 1$ so that the posterior is
%% $p(\theta)(\prod_{i=1}^N p(x_i\mid \theta))^{1/T}$, resulting in the log
%% posterior of 
%% \begin{equation}\label{eq:log_prior_temp}
%%     \log p(\theta \mid x_1,\ldots,x_N) \approx \log p(\theta) + \frac{1}{T}\sum_{i=1}^N\log p(x_i \mid \theta).
%% \end{equation}
%% which has the extra $1/T$ to decrease the scale factor, so the variance scales
%% as $1/T^2$.

@@

%% \begin{figure}[t]
%%   \centering
%%   \includegraphics[width=1\linewidth]{contour_v2}
%%   \caption{The posterior distribution, with $N=10000$ and temperature $T$ varying.}
%%   \label{fig:contour2}
%% \end{figure}

%% To gain intuition on the shape of the posterior, Figure~\ref{fig:contour2} shows
%% simulated contour plots of the posterior based on $N=10000$ data points, and
%% temperature varying.  The (non-tempered) posterior is a multimodal distribution
%% with modes at $(0,1)$ and $(1,-1)$. A larger $T$ implies a flatter posterior,
%% which can be useful for parameter exploration so that the samples do not get
%% trapped in a mode.
